---
title: "Weighted Consensus Segmentation in R"
author: "Halima Saker, Rainer Machne, Jörg Fallmann, Dougie Murray,
Ahmad M. Shahin, Peter F. Stadler"
date: "`r format(Sys.time(), '%d %m %Y')`"
output:
    bookdown::html_document2:
        base_format: rmarkdown::html_vignette
        toc: true
        toc_depth: 2
        fig_caption: true
bibliography: consseg.bib
vignette: >
  %\VignetteIndexEntry{Weighted Consensus Segmentation in R}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, global.par=TRUE,
                      fig.path = ".", fig.pos = 'h',
                      fig.height = 2.7, fig.width = 4, fig.align = "center")
knitr::opts_knit$set(global.par = TRUE)

## redo benchmarking or use existing figure?
REDOLONG <- FALSE
times <- 100
```

```{r, include=FALSE}
par(mai=c(.5,.5,.3,.5),mgp=c(1.2,.3,0),tcl=-.25, cex.main=.75)
```

The R package `consseg` performs

`compileEquation` and `evaluateEquation`,

`plot_breakpoints`, 

R implementation of a dynamic programming algorithm to compute
the consensus of alternative segmentations

# Theory {#theory}
\def\coloneqq{:=}


## Background

A segmentation $\mathbf{S}$ of the interval $[1,n]$ is a partition of
$[1,n]$ into intervals. A consensus segmentation $\mathbf{C}$ for a set
$\{\mathbf{S}_q| 1\le q\le m\}$ of segmentations on $[1,n]$ is a
segmentation that minimizes a weighted sum of distances 
\begin{equation}
f(\mathbf{C}) \coloneqq \sum_{q=1}^m w_q D(\mathbf{C},\mathbf{S}_q)
\end{equation}

We write by $\mathbf{\hat S}$ the \emph{union segmentation}, i.e., the
segmentation that contains of all intersections of the input segments and
thus all input breakpoints. A large class of distances between
segmentations (and partitions in general) can be expressed in the form 
\begin{equation}
  D(\mathbf{S'},\mathbf{S''}) =
  E(\mathbf{S'})+E(\mathbf{S''})-2E(\mathbf{S'}\wedge\mathbf{S''})
\end{equation}
where $E(\mathbf{S})=\sum_{B\in\mathbf{S}} \e(B)$ is the sum of so-called
potentials $\e(B)$ that evaluate the segments $B$ that make up the
segmentation $\mathbf{S}$. In practise, $\e(B)$ is implemented as a function
of the length $z\coloneqq|B|/n$ of the segment $B$ only. Typical choice for the
potential are the \emph{negentropy} $\e(z)= z\ln z$ or power potentials
of the form $z^{\alpha+1}$ with $\alpha>1$. 

## Recursion and Backtracking

For consensus segmentations of the form introduced above, the minimization
problem for $f(\mathbf{C})$ can be solved by dynamic programming. Denote by
$F_k$ the minimal value of $f(\mathbf{C})$ restricted to interval $[1,k]$.
The additive structure of $D(\mathbf{S'},\mathbf{S''})$ in terms of the
potential terms for the individual segments implies that $F_k$ satisfies 
the recursion 
\begin{equation}
  F_k = \min_{j<k} F_j + \left(\Delta([j+1,k]) + F_j \right)
  (\#eq:recursion)
\end{equation}
where the score contribution $\Delta(A)$ for the segment $A=[j+1,k]$ of the
censensus segmentation can be written as 
\begin{equation}
\Delta(A) =
  \e(A) - 2\sum_{q=1}^m w_q \sum_{\substack{B\in \mathbf{S}_q\\
                                 B\cap A\ne\emptyset}}   \e(A\cap B) 
(\#eq:DeltaA)
\end{equation}
Backtracking is performed by recording, for each $k$, the value
$J(k)\coloneqq j$ for the minimum in (\#eq:recursion) is attained.
Starting from $k=n$, one obtains the optimal segmentation as segments
$[J(k)+1,k]$ and setting $k\leftarrow J(k)$ in each step.

## Efficient computation of $\Delta(A)$ 

As defined in (\#eq:DeltaA), evaluation of $\Delta(A)$ requires $O(n^3 m)$
time. This is reduced to $O(n^2m)$ computing the terms recursively in terms
of four auxiliary one-dimensional arrays. In the following we write
$\min B$ and $\max B$ for the mininum and maximum element of a segment $B$.
The first two arrays contain the potential contributions of all segments
that end no later than position $i$ and that begin not later than position
$i$, respectively.
\begin{equation} 
  \delta_{<}(i) \coloneqq \sum_{q=1}^m
  w_q \sum_{\substack{B\in \mathbf{S}_q \\ \max B \le i}} \e(B)
  \qquad\text{and}\qquad
  \delta_{\le}(i) \coloneqq \sum_{q=1}^m
  w_q \sum_{\substack{B\in \mathbf{S}_q \\ \min B\le i}} \e(B)
\end{equation}
Thus $\delta_{<}(k)-\delta_{\le}(j)$ contains the contribution of all
segments that lie within $[j+1,k]$, with the exception of any segments that
straddle the interval, i.e., that begin no later than $j$ and end only
after $k$. These subtracted by $\delta_{\le}(j)$ and need to be corrected
for. The second pair of auxiliary arrays handle the contribution of in
segments $B$ that straddle position $i$, accounting for partial segments
$B$ up to $i$ and from $i$ onwards, respectively:
\begin{equation} 
  \delta^{\cap}_{<}(i) \coloneqq \sum_{q=1}^m w_q 
  \sum_{\substack{B\in \mathbf{S}_q \\ i\in B, i\ne\max B}}
  \e(B\cap[1,i])
  \qquad\text{and}\qquad
  \delta^{\cap}_{>}(i) \coloneqq \sum_{q=1}^m w_q 
  \sum_{\substack{B\in \mathbf{S}_q \\ i\in B, i\ne\min B}}
  \e(B\cap[i,n])
\end{equation}
Given an interval $[j+1,k]$, the sum $\delta_{<}(k)+\delta_{>}(j+1)$
accounts for the intersection for contributions of segments $B$ with
$B\cap A\ne\emptyset$ and $B\not\subseteq A$. Again intervals $B$
containing $[j,k+1]$ contribute to these terms. These corrections are
summarized in
\begin{equation}
  \delta^*(i',i'') \coloneqq \sum_{q=1}^m w_q
  \sum_{\substack{B\in \mathbf{S}_q \\ i',i''\in B \\
      \min B< i'\le i''<\max B}}
      \big( \e(B) + \e([i',i'']) - \e(B\cap[1,i'']) - \e(B\cap[i',n])\big)\,,
\end{equation}
which can be computed in $O(m)$ time since there is at most one interval
$B$ per input segmentation $\mathbf{S}_q$ that contains both $i'$ and
$i''$. With the help of the auxiliary arrays. The auxiliary arrays
$\delta_{<}$ and $\delta_{\le}$ themselves can be computed stepwisely,
with incremental effort $O(m)$ for each position, while 
$\delta^{\cap}_{<}(i)$, $\delta^{\cap}_{>}(i)$ are sums of at most
$m$ terms, causing a total setup cost of $O(n m)$. We have 
\begin{equation}
  \Delta([j+1,k]) = \e([j+1,k])
  -2 \big(
    \delta_{<}(k) - \delta_{\le}(j) +
    \delta^{\cap}_{<}(k)+ \delta^{\cap}_{>}(j+1) + \delta^*(j+1,k)
  \big)
(\#eq:Delta)
\end{equation}
which can now be evaluated for given $j$ and $k$ in $O(m)$ time. Note that
each combination of $j$ and $k$ appears only once in the course of the
recursion for $F_k$. Thus there is no point in storing the quadratically
many correction terms $\delta^*(j+1,k)$ or the final scores.

## Segment Length Restrictions

For a large class of potential functions, including the power potentials
$\e(z)=z^{1+\alpha}$ with $\alpha>1$, it can be shown that no segment in
the consensus $\mathbf{C}$ can be longer than twice the length $L$ of
longest input segment. This can be used to restrict the range of the
recursion to $|k-j+1|\le 2L$. If such \emph{a priori} bounds on segment
length are known, they can be used to speed up the computation, reducing
the effort from $O(n^2 m)$ to $O(n m L)$. 

# Usage

## Basic


```{r conssegdemo, fig.cap=".\\label{fig:consseg}"}
library(consseg)

n <- 50# 5000 #SEQUENCE LENGTH
M <- 10 # number segmentations (breakpoint lists)
l <- 4# # average number of segments

set.seed(1) # for constant results
b <- random_breakpoints(m=M,n=n,lambda=l)

w <- rep(1/M, M)

e <- "(L/n)*log(L/n)"
cons <- consensus(b, n=n, w=w, e=e, verb=1)

plot_breaklist(b)
abline(v=cons, col="#0000FFCC", lwd=2)

```


## Custom Potential Functions {#generic}

```{r potential, fig.cap=".\\label{fig:consseg}"}

## compile a novel potential function
e <- "(L/n)*log(L/n)"
ec <- compileEquation(e)

## scan over L
res <- rep(NA, 500)
for ( L in 1:length(res) )
    res[L] <- evaluateEquation(e=ec, L=L, n=500)

plot(1:length(res),res, xlab="L", ylab=e)
```

## Real Data Use Case

calculate consensus from segmenTier data

# Benchmarking


# Dynamic Programming in base R {#appi}

## Direct Implementation in Base R

```{r, eval=TRUE, echo=TRUE}
## RECURSION

## POTENTIAL FUNCTION

## backtracing
```

## Incremental Calculation




# References
