---
title: "Weighted Consensus Segmentation in R"
author: "Halima Saker, Rainer Machne, Jörg Fallmann, Ahmad M.\ Shahin, Peter F. Stadler"
date: "`r format(Sys.time(), '%d %m %Y')`"
output:
    bookdown::html_document2:
        base_format: rmarkdown::html_vignette
        toc: true
        toc_depth: 2
        fig_caption: true
bibliography: consseg.bib
vignette: >
  %\VignetteIndexEntry{Weighted Consensus Segmentation in R}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
  \usepackage{mathtools}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, global.par=TRUE,
                      fig.path = ".", fig.pos = 'h',
                      fig.height = 2.7, fig.width = 4, fig.align = "center")
knitr::opts_knit$set(global.par = TRUE)

## redo benchmarking or use existing figure?
REDOLONG <- FALSE
times <- 100
```

```{r, include=FALSE}
par(mai=c(.5,.5,.3,.5),mgp=c(1.2,.3,0),tcl=-.25, cex.main=.75)
```

$$
\newcommand{\e}{\mathfrak{e}}
\newcommand{\coloneqq}{:=} %TODO: make mathtools work
$$

The R package `consseg` performs

`compileEquation` and `evaluateEquation`,

`plot_breakpoints`, 

R implementation for testing and development ...

# Theory {#theory}

## Recursion

$$
\begin{equation}
\Delta(A)
(\#eq:DeltaA)
\end{equation}
$$

$$
\begin{equation}
F_k = \min_{j<k} \left(\Delta([j+1,k]) + F_j \right)
(\#eq:recursion)
\end{equation}
$$



## Potential Functions

## Backtracking 

## Segment Length Restrictions


# Usage

## Basic


```{r conssegdemo, fig.cap=".\\label{fig:consseg}"}
library(ConsSeg)

n <- 50# 5000 #SEQUENCE LENGTH
M <- 10 # number segmentations (breakpoint lists)
l <- 4# # average number of segments

set.seed(1) # for constant results
b <- random_breakpoints(m=M,n=n,lambda=l)

w <- rep(1/M, M)

e <- "(L/n)*log(L/n)"
cons <- consensus(b, n=n, w=w, e=e, verb=1)

plot_breaklist(b)
abline(v=cons, col="#0000FFCC", lwd=2)

```


## Custom Potential Functions {#generic}

```{r potential, fig.cap=".\\label{fig:consseg}"}

## compile a novel potential function
e <- "(L/n)*log(L/n)"
ec <- compileEquation(e)

## scan over L
res <- rep(NA, 500)
for ( L in 1:length(res) )
    res[L] <- evaluateEquation(e=ec, L=L, n=500)

plot(1:length(res),res, xlab="L", ylab=e)
```

## Real Data Use Case

calculate consensus from segmenTier data

# Benchmarking


# Dynamic Programming in base R {#appi}

## Direct Implementation in Base R

```{r, eval=TRUE, echo=TRUE}
## RECURSION

## POTENTIAL FUNCTION

## backtracing
```

## Incremental Calculation

$$
\begin{equation} 
  \delta_{<}(i) \coloneqq \sum_{q=1}^m
  w_q \sum_{B\in \mathbf{S}_q \atop \max B \le i} \e(B)
  \qquad\text{and}\qquad
  \delta_{\le}(i) \coloneqq \sum_{q=1}^m
  w_q \sum_{B\in \mathbf{S}_q \atop \min B\le i} \e(B)
\end{equation}
$$

TODO: make the following work, perhaps its atop?

$$
\begin{equation} 
  \delta^{\cap}_{<}(i) := \sum_{q=1}^m w_q 
  \sum_{B\in \mathbf{S}_q \atop i\in B, i\ne\max B}

  \e(B_{\le i})
  \qquad\text{and}\qquad
  \delta^{\cap}_{>}(i) := \sum_{q=1}^m w_q 
  \sum_{B\in \mathbf{S}_q \atop i\in B, i\ne\min B}

  \e(B_{\ge i})
(\#eq:deltacup)
\end{equation}
$$



$$
\begin{equation}
  \Delta([j+1,k]) = \e([j+1,k])
  -2 \big(
    \delta_{<}(k) - \delta_{\le}(j) +
    \delta^{\cap}_{<}(k)+ \delta^{\cap}_{>}(j+1) + \delta^*(j+1,k)
  \big)
(\#eq:Delta)
\end{equation}
$$


# References
