---
title: "Weighted Consensus Segmentation in R"
author: "Halima Saker, Rainer Machne, J{\"o}rg Fallmann, Ahmad M.\ Shahin, Peter F. Stadler"
date: "`r format(Sys.time(), '%d %m %Y')`"
output:
    bookdown::html_document2:
        base_format: rmarkdown::html_vignette
        toc: true
        toc_depth: 2
        fig_caption: true
bibliography: consseg.bib
vignette: >
  %\VignetteIndexEntry{Weighted Consensus Segmentation in R}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, global.par=TRUE,
                      fig.path = ".", fig.pos = 'h',
                      fig.height = 2.7, fig.width = 4, fig.align = "center")
knitr::opts_knit$set(global.par = TRUE)

## redo benchmarking or use existing figure?
REDOLONG <- FALSE
times <- 100
```

```{r, include=FALSE}
par(mai=c(.5,.5,.3,.5),mgp=c(1.2,.3,0),tcl=-.25, cex.main=.75)
```


The R package `consseg` performs

`compileEquation` and `evaluateEquation`,

`plot_breakpoints`, 

R implementation for testing and development ...

# Theory {#theory}

## Recursion

## Potential Functions

## Backtracing 

## Segment Length Restrictions


# Usage

## Basic


```{r conssegdemo, fig.cap=".\\label{fig:consseg}"}
library(ConsSeg)

n <- 50# 5000 #SEQUENCE LENGTH
M <- 10 # number segmentations (breakpoint lists)
l <- 4# # average number of segments

set.seed(1) # for constant results
b <- random_breakpoints(m=M,n=n,lambda=l)

w <- rep(1/M, M)

e <- "(L/n)*log(L/n)"
cons <- consensus(b, n=n, w=w, e=e, verb=1)

plot_breaklist(b)
abline(v=cons, col="#0000FFCC", lwd=2)

```


## Custom Potential Functions {#generic}

```{r potential, fig.cap=".\\label{fig:consseg}"}

## compile a novel potential function
e <- "(L/n)*log(L/n)"
ec <- compileEquation(e)

## scan over L
res <- rep(NA, 500)
for ( L in 1:length(res) )
    res[L] <- evaluateEquation(e=ec, L=L, n=500)

plot(1:length(res),res, xlab="L", ylab=e)
```

## Real Data Use Case

calculate consensus from segmenTier data


# Dynamic Programming in base R {#appi}



```{r, eval=TRUE, echo=TRUE}
## RECURSION

## POTENTIAL FUNCTION

## backtracing
```

# Benchmarking


# References
